```{r echo = FALSE}
library(distill)
```

---
title: "Bayesian inference in computational intelligence"
author: 
  - name: "Rodrigo Pereira Cruz"
    email: pereirarodrigocs@gmail.com
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

This is a notebook used for studying Bayesian inference in computational intelligence, which is focused on its theoretical applications. The R programming language was chosen for the practical part of this project, given its orientation towards statistical analysis, ease of use and modular nature makes it well suited for this kind of task.  

---

# 1. Introduction

***

Bayes' theorem is a fundamental part of probability and statistics. It is deeply tied to conditional probability, often expressed as $P(A | B)$ and read as the probability of A given B. Bayesian statistics relies on having prior probabilities, which constitutes the currently known information, and calculating posterior probabilities based on the prior and likelihood. This is a constant process, where the prior probabilities are constantly updated. Bayes' theorem is usually defined as:

<br>

$$P(A | B) = \frac{P(B | A) \space P(A)}{P(B)}$$

<br>

Where:

<br>

* $P(A | B)$ = posterior probability;
* $P(B | A)$ = likelihood;
* $P(A)$ = prior probability;
* $P(B)$ = marginal probability.

<br>

Oftentimes, this definition is expanded upon when $P(B)$ isn't clearly defined:

<br>

$$\begin{align}
    P(A | B) &= \frac{P(B | A) \space P(A)}{P(B)} \\
             & = \space \frac{P(B | A) \space P(A)}{P(B | A) \space P(A) + P(B | \neg A) \space P(\neg A)}
\end{align}$$

<br>

When it comes to computing, Bayes' theorem is already popular in the field of machine learning, where it's known as the naive Bayes algorithm and is used as a classifier. A simple usage can be seen below:

```{r}

# A simple use of the naive Bayes model for classification
# The model efficiently classifies 3 types of iris flowers: setosa, versicolor and virginica

library(e1071)
library(caret)

data("iris")

iris_dataset <- as.data.frame(iris)
bayes <- naiveBayes(Species ~., data = iris_dataset)

# We can check the model by typing "bayes"
bayes

bayes_predictions <- predict(bayes, iris_dataset)

table(bayes_predictions, iris_dataset$Species)

confusionMatrix(iris$Species, bayes_predictions)
```
